---
title: "Importance sampling for Bootstrapping of posterior means"
author: "Topi Paananen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Importance sampling for Bootstrapping of posterior means}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r SETTINGS-knitr, include=FALSE}
stopifnot(require(knitr))
options(width = 90)
opts_chunk$set(
  comment = NA,
  message = FALSE,
  warning = FALSE,
  # eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  dev = "png",
  dpi = 150,
  fig.asp = 0.8,
  fig.width = 5,
  out.width = "60%",
  fig.align = "center"
)
```

## Introduction

In this case study, we will study generic methods to detect
misspecification of Bayesian models. We will use the probabilistic
programming framework Stan to fit our models.  As a tool for detecting
misspecification, we will look at the distribution of posterior means
when Bootstrapping the data. For a correctly specified model, as the
number of observations becomes large enough, the posterior
distribution of the model and the distribution of Bootstrapped
posterior means should match each other.

In order to compare the model posterior to the distribution of the
Bootstrap means, we need a large number of Bootstrap repetitions.
Because refitting the model to a large number of resampled data sets
is expensive, we will only fit our model to the original data set and
use importance sampling to speed up the Bootstrap procedure. This
saves computation time but can give rise to errors when there is a
large mismatch between the original posterior and the Bootstrap
posteriors.  To solve this issue, we use importance weighted moment
matching (IWMM) which is a generic adaptive importance sampling
algorithm (Paananen et al. 2020).

## Setup

We will load __rstan__ for fitting our models, and the
[_iwmm package_](https://github.com/topipa/iwmm) for
the importance weighted moment matching.


```{r load, message=FALSE}
library("rstan")
library("iwmm")
seed <- 48571056
set.seed(seed)
```

## Example 1: Correct model - Gaussian data and Gaussian model

### Coding the Stan model

```{r stancode}
stancode <- "data {
  int<lower=0> N;
  vector[N] x;
}
parameters {
  real mu;
  real logsigma;
}
transformed parameters {
  real<lower=0> sigma;
  sigma = exp(logsigma);
}
model {
  x ~ normal(mu, sigma);
}
generated quantities {
  vector[N] log_lik;
  vector[N] xpred;
  for (n in 1:N)
  {
    log_lik[n] = normal_lpdf(x[n] | mu, sigma);
    xpred[n] = normal_rng(mu, sigma);
  }
}"
```

For the Bootstrap importance sampling, we need the log-likelihood for
each observation.  Thus, following the usual approach recommended in
[_Writing Stan programs for use with the loo
package_](http://mc-stan.org/loo/articles/loo2-with-rstan.html), we
compute the log-likelihood for each observation in the `generated
quantities` block of the Stan program.

### Fitting the model with RStan

First we simulate Gaussian data and fit the correctly specified model with
Stan using the rstan package:

```{r modelfit1, message=FALSE, echo=T, results="hide"}
stanmodel <- stan_model(model_code = stancode)

# generate  data
n <- as.integer(100)
x <- rnorm(n = n)

standata <- list(N = n, x = x)
stanfit <- sampling(stanmodel, data = standata, refresh = 0)

# extract posterior and log-likelihood draws
post <- as.data.frame(stanfit)[, 1:2]
S <- nrow(post)
ll <- as.matrix(stanfit, pars = "log_lik")
```

### Compute Bootstrap means using importance sampling

Now let us resample our simulated data in a Bootstrap fashion, and use importance
sampling to compute the Bootstrap posterior means. To diagnose the reliability of importance
sampling, we compute the Pareto k diagnostics (See Vehtari et al. (2017) for details).

```{r naiveIS, warning=FALSE}
bs_S <- 5

# compute Bootstrap weights
obs_ws <- apply(
  matrix(seq(bs_S)), 1,
  function(x) c(rmultinom(1, n, prob = rep(1, n) / n))
)

bs_postmeans_is <- matrix(0, bs_S, 2)
ks <- rep(0, bs_S)

iw_list <- lapply(X = seq(bs_S), FUN = function(i) {
  obs_w_matrix <- matrix(obs_ws[, i], S, n, byrow = TRUE)
  lw <- rowSums((obs_w_matrix - 1) * ll)
  lw <- lw - matrixStats::logSumExp(lw)
  khat <- posterior::pareto_khat(lw, are_log_weights = TRUE, tail = "right")
  c(colSums(exp(lw) * post), khat)
})

iw_list <- array(as.numeric(unlist(iw_list)), c(3, bs_S))
bs_postmeans_is <- data.frame("mu" = iw_list[1, ], "logsigma" = iw_list[2, ])
ks <- iw_list[3, ]
```

Let us plot the Pareto k diagnostics to see if the importance sampling
accuracy can be trusted:

```{r plotk1}
plot(sort(ks), pch = 20)
```

We notice that there are several Pareto k values that are large (>
0.7), indicating that the Bootstrap means can be biased.

### Compute Bootstrap means using IWMM

In order to get accurate results with importance sampling, we will use the
iwmm algorithm. This will require more computations compared to simple
importance sampling, but is still significantly cheaper than
refitting the model for each resampled data set.

We will use the `moment_match` function from the __iwmm__ package.
There is a separate version of the function for `stanfit` objects.
The function requires two arguments: 1) A `stanfit` object, and 2)
a function `log_ratio_fun` that computes the ratio of the target density
(the Bootstrap posterior) and the proposal density (the standard posterior).
For `log_ratio_fun`, we give a third argument that represents the
weights of each observation, which in this case represent the number of
times each observation is included in the Bootstrap sample.

```{r helper functions, warning = FALSE}
log_lik_stanfit <- function(fit, upars, parameter_name = "log_lik",
                            ...) {
  ll <- as.matrix(fit, pars = parameter_name)
  S <- nrow(upars)
  n <- ncol(ll)
  out <- matrix(0, S, n)
  for (s in seq_len(S)) {
    out[s, ] <- rstan::constrain_pars(
      fit,
      upars = upars[s, ]
    )[[parameter_name]]
  }
  out
}

ratio_density <- function(draws, fit, obs_weights, ...) {
  log_lik <- log_lik_stanfit(fit, draws)
  colSums((obs_weights - 1) * t(log_lik))
}
```

```{r iwmm1, warning = FALSE}
bs_postmeans_is_mm <- matrix(0, bs_S, 2)
ks_mm <- rep(0, bs_S)


iw_list <- lapply(X = seq(bs_S), FUN = function(i) {
  iw <- moment_match(
    stanfit,
    log_ratio_fun = ratio_density,
    obs_weights = obs_ws[, i]
  )
  c(matrixStats::colWeightedMeans(
    posterior::as_draws_matrix(iw$draws)[, c("mu", "logsigma")],
    exp(iw$log_weights)
  ), iw$pareto_k)
})
iw_list <- array(as.numeric(unlist(iw_list)), c(3, bs_S))

bs_postmeans_is_mm <- data.frame("mu" = iw_list[1, ], "logsigma" = iw_list[2, ])
ks_mm <- iw_list[3, ]
```

Let us check the Pareto k diagnostic values :

```{r plotk2}
plot(sort(ks),
  pch = 20,
  ylim = c(min(c(ks, ks_mm)), max(c(ks, ks_mm))),
  col = "black",
  xlab = "Bootstrap index",
  ylab = "Pareto k",
  main = "Pareto k diagnostics"
)
points(ks_mm[order(ks)], col = "green", pch = 20)
points(c(-1, 1000), c(0.5, 0.5), type = "l")
legend("topleft", legend = c("IS", "IWMM"), col = c("black", "green"), pch = 20)
```

Now all values are below 0.5, so the Bootstrap means should be
accurate.  Let us plot both the original posterior samples and the
Bootstrap posterior means:

```{r plotdraws1}
hist(post$mu,
  col = rgb(0, 0, 1, 1 / 4),
  freq = FALSE,
  ylim = c(0, 25),
  xlab = "mu",
  ylab = "count",
  main = "Posterior of mu"
)
hist(bs_postmeans_is_mm$mu, col = rgb(1, 0, 0, 1 / 4), add = TRUE, freq = FALSE)
legend("topright",
  legend = c("model posterior", "Bootstrap posterior means"),
  fill = c(rgb(0, 0, 1, 1 / 4), rgb(1, 0, 0, 1 / 4))
)
```

```{r plotdraws2}

hist(post$logsigma,
  col = rgb(0, 0, 1, 1 / 4),
  freq = FALSE,
  ylim = c(0, 50),
  xlab = "log(sigma)",
  ylab = "count",
  main = "Posterior of log(sigma)",
)
hist(bs_postmeans_is_mm$logsigma,
  col = rgb(1, 0, 0, 1 / 4),
  add = TRUE, freq = FALSE,
#  breaks = seq(-0.1, 0.15, length.out = 30)
)
legend("topright",
  legend = c("model posterior", "Bootstrap posterior means"),
  fill = c(rgb(0, 0, 1, 1 / 4), rgb(1, 0, 0, 1 / 4))
)
```

As we expect from a correctly specified model, the distribution of the
Bootstrap posterior means matches the posterior distribution of the
original data.  Let us next see what happens if the model is clearly
misspecified.

## Example 2: Misspecified model - $t$-distributed data and Gaussian model

```{r modelfit2, message=FALSE}
stanmodel <- stan_model(model_code = stancode)

# generate  data
n <- as.integer(1000)
x <- rt(n = n, df = 3)

standata <- list(N = n, x = x)
stanfit <- sampling(stanmodel, data = standata, refresh = 0)

# extract posterior and log-likelihood draws
post <- as.data.frame(stanfit)[, 1:2]
S <- nrow(post)
ll <- as.matrix(stanfit, pars = "log_lik")
```


### Compute Bootstrap means using IWMM

```{r iwmm2, warning = FALSE}
bs_postmeans_is_mm <- matrix(0, bs_S, 2)
ks_mm <- rep(0, bs_S)

iw_list <- lapply(X = seq(bs_S), FUN = function(i) {
  iw <- moment_match(
    stanfit,
    log_ratio_fun = ratio_density,
    obs_weights = obs_ws[, i]
  )
  c(matrixStats::colWeightedMeans(
    posterior::as_draws_matrix(iw$draws)[, c("mu", "logsigma")],
    exp(iw$log_weights)
  ), iw$pareto_k)
})
iw_list <- array(as.numeric(unlist(iw_list)), c(3, bs_S))

bs_postmeans_is_mm <- data.frame("mu" = iw_list[1, ], "logsigma" = iw_list[2, ])
ks_mm <- iw_list[3, ]
```


Let us check the Pareto k diagnostic values :

```{r plotk3}
plot(sort(ks_mm),
  pch = 20,
  ylim = c(min(ks_mm), max(ks_mm)),
  col = "green",
  xlab = "Bootstrap index",
  ylab = "Pareto k",
  main = "Pareto k diagnostics"
)
# points(ks_mm[order(ks)], col = "green", pch = 20)
points(c(-1, 1000), c(0.5, 0.5), type = "l")
legend("topleft", legend = c("IS", "IWMM"), col = c("black", "green"), pch = 20)
```

All diagnostic values are below 0.5, so the Bootstrap means should be accurate.
Let us plot both the original posterior samples and the Bootstrap posterior means:

```{r plotdraws3}
hist(post$mu,
  col = rgb(0, 0, 1, 1 / 4),
  freq = FALSE,
  ylim = c(0, 25),
  xlab = "mu",
  ylab = "count",
  main = "Posterior of mu"
)
hist(bs_postmeans_is_mm$mu, col = rgb(1, 0, 0, 1 / 4), add = TRUE, freq = FALSE)
legend("topright",
  legend = c("model posterior", "Bootstrap posterior means"),
  fill = c(rgb(0, 0, 1, 1 / 4), rgb(1, 0, 0, 1 / 4))
)
```

```{r plotdraws4}
hist(post$logsigma,
  col = rgb(0, 0, 1, 1 / 4),
  freq = FALSE,
  ylim = c(0, 50),
  xlab = "log(sigma)",
  ylab = "count",
  main = "Posterior of log(sigma)",
#  breaks = seq(0.2, 1.0, length.out = 30)
)
hist(bs_postmeans_is_mm$logsigma,
  col = rgb(1, 0, 0, 1 / 4),
  add = TRUE, freq = FALSE,
#  breaks = seq(0.2, 1.0, length.out = 30)
)
legend("topright",
  legend = c("model posterior", "Bootstrap posterior means"),
  fill = c(rgb(0, 0, 1, 1 / 4), rgb(1, 0, 0, 1 / 4))
)
```

Now we see that there is a clear difference between the two
distributions.  Because the data is from a $t$-distribution, but we
are modelling it as Gaussian, the model underestimates the uncertainty
about the standard deviation parameter `logsigma`.

We can confirm a similar misspecification using posterior predictive
checks.  Below we plot a density plot of the data together with
multiple data sets simulated from the model.

```{r}
xpred <- extract(stanfit)$xpred
bayesplot::ppc_dens_overlay(x, xpred[1:100, ])
```

## References

Paananen, T., Piironen, J., Bürkner, P.-C., and Vehtari, A.
Implicitly Adaptive Importance Sampling. _Stat Comput_ **31**, 16 (2021).
([paper](https://doi.org/10.1007/s11222-020-09982-2))([code](https://github.com/topipa/iter-mm-paper))

Stan Development Team (2023) _RStan: the R interface to Stan, Version 2.21.8_   https://mc-stan.org

Vehtari, A., Gelman, A., and Gabry, J. (2017). Practical Bayesian model evaluation using leave-one-out cross-validation and WAIC. _Statistics and Computing_. 27(5), 1413--1432. \doi:10.1007/s11222-016-9696-4. Links: [published](http://link.springer.com/article/10.1007\%2Fs11222-016-9696-4) | [arXiv preprint](http://arxiv.org/abs/1507.04544).


